---
title: "Danish Kings"
author: "Marius Henningsen"
date: "2025-01-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator: kings are separated by ","; `antonkings_csv` by "`"

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval= TRUE}

library(tidyverse)

# FILL IN THE CODE BELOW and review the outputs

kings1 <- read.csv("data/kings.csv")

kings2 <- read_csv("data/kings.csv")

kings3 <- read.csv2("data/kings.csv")

kings4 <- read_csv2("data/kings.csv")



kings1
glimpse(kings1)
class(kings1)
head(kings1)

kings2
glimpse(kings2)
class(kings2)
head(kings1)

kings3
glimpse(kings3)
class(kings3)
head(kings3)

kings4
glimpse(kings4)
class(kings4)
head(kings4)

```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object

read_csv() and read_csv2() are `tidyverse` functions

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

you get this result: "spec_tbl_df" "tbl_df"      "tbl"         "data.frame" 

3. How many columns does the object have when created with these different functions?

In the cases of read.csv and read_csv the object creates 11 columns
In the cases of read.csv2 and read_csv2 the object creates 1 column

4. Show the dataset so that we can see how R interprets each column



```{r eval = TRUE}

# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- read_csv("data/kings.csv", na = c("NULL","")) 
# this is the right function for my dataset

class(kings)


glimpse(kings)


head(kings)



```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE

kings <- kings %>% 
  mutate(Start_year = as.numeric(Start_year), End_year = as.numeric(End_year))
  
kings <- kings %>% 
  mutate(Duration = End_year - Start_year) 
  


top_3 <- sort(kings$Duration, decreasing = TRUE)[1:3]

top_3


```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
# YOUR CODE

library(tidyverse)

# Tibble way

avg_reign <- kings %>% 
  summarize(mean_duration = mean(Duration, na.rm = TRUE))

# Vector way

avg_reign_vector <- mean(kings$Duration, na.rm = TRUE)

print(avg_reign)


```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
# YOUR CODE


#1 Finds the kings who enjoyed a longer-than-average duration of reign
long_reigns <- kings %>% 
  filter(Duration > avg_reign_vector)
long_reigns

#2 Shows which kings enjoyed a longer-than-average duration of reign
long_reigns %>% 
  select(Name, Duration) %>% 
  arrange(desc(Duration))

# 5. Counts how many kings enjoyed a longer-than-average duration of reign
long_reigns %>% 
  count()

```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# YOUR CODE

top_3_monarchs <- kings %>% 
  arrange(desc(Duration)) %>%   
  slice(1:3) %>%                
  mutate(
    Days_Basic = Duration * 365,
    # To count transition/leap year we add 0.25 to the 365 days, because every fourth year is a leap year
    Days_Leap = Duration * 365.25 
  )

print(top_3_monarchs %>% select(Name, Duration, Days_Leap))

```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE

kings <- kings |>
  mutate(Midyear = Start_year + (End_year - Start_year) / 2)

ggplot(kings, aes(x = Midyear, y = Duration)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "Duration of reign of Danish monarchs through time",
    x = "Midyear of reign",
    y = "Duration of reign (years)"
  ) +
  theme_bw() +
  theme(text = element_text(size = 14))

```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


